Computational methods in landscape ecology

### 2024-06-02 Sections/spatial-patterns.qmd
starts vectormetrics section

## Vector metrics
<!-- Nowosad, Hesselbarth -->

an intro to the section
what is similar to landscape metrics

add an info about shape metrics
@basaranerPerformanceShapeIndices2017

https://github.com/r-spatialecology/vectormetrics

<!--misc-->
<!-- the need for open-source software in landscape ecology -->
<!-- every proposed method should be open-source and available to the community -->

>>>>

## Vector metrics
<!-- Nowosad, Hesselbarth -->
<!-- jn: ask T. Matuszek for review of this section later -->

As already mentioned, the initial computaional approaches in landscape ecology were based on raster data.
However, the decision on the used data model was not theorethical, but rather practical as at the time the most comonnly used software was raster-based, such as GRASS GIS.
In fact, early studies were based on polygonal vector data that was later converted to raster data for the analysis [@costanzaDescribingAnalyzingLandscape2019]<!--jn: we could also cite actual examples from that paper, but the list of the references would be long-->
<!-- jn: maybe we could also mention that it is still the case -->

The spatial vector data model is based on the representation of the Earth's surface as a set of points, lines, and polygons, where each feature is defined by its geometry and attributes.
Here, we will focus on the polygonal vector data, which is the most common in landscape ecology.
The polygons are discrete areas, with well-defined boundaries.
Thus, converting the polygons to raster data (*rasterization*) is a lossy process, where the information about the intricates of the polygon shapes is lost.
Moreover, the rasterization process largely depends on the decided resolution of the raster outcome.

Most landscape ecology studies are still based on raster data, due to the familarity with the techiniques, the availability of the data, and the availability of the software.
For example, most land cover maps (as LULC products are often related to undelining remote sensing products), and all on the global scale, are provided as rasters. 
At the same time, nowadays, vector data of land cover on regional or local scales is also available and most geospatial software support also vector data.

As hundreds of metrics have been already developed for raster data, the most straightforward approach is to reimplement the same metrics for vector data.
@yaoVecLIFrameworkCalculating2022 showed that this is possible for most of the metrics, with only several exceptions of metrics (of the shape and aggregation group), which parameters are related to the raster-format structure.
Their results highlighted that about 80% of the raster and vector-based metrics are highly correlated (for their study area), and the higher the raster resolution, the stronger the correlation is.

<!-- metrics from other fields -->
<!-- add an info about shape metrics -->
<!-- @basaranerPerformanceShapeIndices2017 -->

<!-- https://github.com/r-spatialecology/vectormetrics -->

<!-- vectormetrics limitations?? 
surely, large data performance is one of them
(e.g., could we do landscapemetrics vs vectormetrics benchmarking?) -->

<!-- polygon vs multipolygon technical distinction -->
